---
title: "Day 1: Intro to R & tidyverse Supplementary Notebook"
author: Josh Shapiro
date: 2021-11-01
output:   
  html_notebook: 
    toc: true
    toc_float: true
---


## Introduction

This notebook addresses some questions that came up after the first day of the CCDL workshop.

- [Installing packages](#installing-packages)
- [`dplyr` join functions](#dplyr-join-functions)

### How to use this notebook: 

While it is fine to just read this page as is, if you want to get the Rmarkdown notebook itself to work with, you can follow the following steps:

1. Click the `Code` button in the upper right of this page and choose `Download Rmd` to save this notebook (`2021-09-20_muddiest_points.Rmd`) to your computer.
If you have RStudio installed on your computer, you can then open the file directly.

Otherwise you can add it to the CCDL RStudio server to work with it there with the following additional steps:

2. Navigate to RStudio server.  
3. In the `File` pane, navigate to `training-modules/intro-to-R-tidyverse`.     
4. Click the `Upload` button in the `File` pane.   
5. Click `Choose file` and find this `Rmd` file on your computer, then click `OK`  


Now you can open up this file and run it on the Server. 

## Installing packages

We do not explicitly cover package installation during this training, mostly for time reasons, but also because we want to make sure everything is there and ready to go!

But of course when you go to work on your own machine, some R packages may not be installed, so you may need to know how to install them!
Usually, this is pretty straightforward, and we will cover the most common situations here.

There are two main sources for R packages that we use, [CRAN](https://cran.r-project.org/) and [Bioconductor](https://bioconductor.org/).
Installation of packages from the two sources is slightly different.

### CRAN packages

Most general use packages (not biology-specific) are available through CRAN, the Comprehensive R Archive Network.
To install a package from CRAN, you use the `install.packages()` function, which will install the requested package(s) as well as any packages they might depend on.
So if I wanted to install (or update) the `svglite` package (which allows `ggplot2::ggsave` to write out SVG files), I would use the following command:

```{r, eval = FALSE}
install.packages("svglite")
```

If I had more than one package to install, I might use a vector of the package names (each enclosed by quotes):

```{r eval = FALSE}
install.packages(c("magrittr", "dplyr", "palmerpenguins"))
```

The first time you use `install.packages()` you may be prompted (in the Console) to choose a "mirror". 
You will need to enter any responses in the console.
Using option `0`, the "cloud", is usually safe, but you can choose another location if you prefer.

If you are using RStudio, there is also an "Install" button in the `Packages` tab, which will prompt you to enter the packages you want to install, but this will only work with CRAN packages.

### Bioconductor packages

To install [Bioconductor](https://bioconductor.org/) packages, you need to use a specialized installation function from the `BiocManager` package.
So the first step is to install that package from CRAN:

```{r eval = FALSE}
install.packages("BiocManager")
```

Once that is installed, you can use `BiocManager::install()` to install any packages that are part of Bioconductor and their dependencies. 
(In fact you could use this always: it will find CRAN packages as well!)

```{r, eval = FALSE}
BiocManager::install("DESeq2")
```

`BiocManager::install()` tends to be a bit more chatty than the base R installation; be on the lookout for prompts that you need to respond to in the console.

## `dplyr` join functions

I briefly talked about join functions, which can be used to merge two data frames together. 
They do this by looking at a column (or columns) to find matching rows, then make a new data frame with all of the columns from both of the original data frames (though usually only one copy of the column that was used for matching).
As there are a variety of ways one might want the matching to proceed, there are a nubmer of different join functions, so we will illustrate a few of them below.

First, let's create a couple of data frames. 
These will share a column called `id`, but have otherwise different columns.
Some of the values in the id column will be the same between the data frames, but some will appear in only one of the two data frames.

```{r}
foo_df <- data.frame(
  id = c("a1", "a2", "a3", "b1", "b2", "b3"),
  foo_value = c(1, 2, 3, 4, 5, 6),
  foo_batch = c(1, 1, 2, 2, 3, 3)
)
foo_df
```
```{r}
bar_df <- data.frame(
  id = c("b3", "b2", "b1", "c1", "c2", "c3"),
  bar_value = c(21, 45, 32, 12, 30, 47),
  bar_batch = c(1, 1, 2, 2, 3, 3)
)
bar_df
```

Notice that the `id` values are not in the same order, so we can't just assume that if we pasted the two data frames side by side we would get the same values.

### Inner join

An inner join (`dplyr::inner_join()`) will find the matches between the two data frames in the specified columns and create a new table with only the shared rows.
We give it the two data frames we want to join as the first two arguments, followed by the `by` argument to say which column(s) from each to use for matching.
Here I am matching on the `id` column from each of the two data frames, and I am being very explicit by giving the first table column name and the second column name in quotes, separated by `=`.

```{r}
dplyr::inner_join(foo_df, bar_df, by = c("id" = "id"))
```

If the column names I want to join on are the same, I can also just use the single column name. 
(I could even leave the `by` argument off completely, and let the join function guess, but that is dangerous!)

```{r}
dplyr::inner_join(foo_df, bar_df, by = "id")
```

I can also join on more than one column at a time:
Here I will join both on the `id` columns and the respective `_batch` columns.

```{r}
dplyr::inner_join(foo_df, bar_df, by = c("id" = "id", "foo_batch" = "bar_batch"))
```
Notice that `bar_batch` no longer appears, because it would be redundant with `foo_batch`.

### Left (and right) join

Other joins follow the same syntax, but return different results. 
Left join (`dplyr::left_join()`)will return a row for every value in the first data frame, whether or not a match is found in the second data frame.
If there is no match, the column values that would have come from the second data frame are filled in wiht `NA`:

```{r}
dplyr::left_join(foo_df, bar_df, by = "id")
```

Right join is the same, but keeping all values from the _second_ data frame.

```{r}
dplyr::right_join(foo_df, bar_df, by = "id")
```
### Full join

Last we have full join, which will include _all_ rows from both tables whether or not there is a match, filling in `NA` as necessary.

```{r}
dplyr::full_join(foo_df, bar_df, by = "id")
```

### Other join features

I've really only scratched the surface here, exploring a very simple example. 
There is a lot more to explore, including what happens when a value in the match column(s) appears more than once in one of the data frames.

For more information, we recommend Jenny Bryan's excellent "[join cheatsheet](https://stat545.com/join-cheatsheet.html)". 
(Don't let the name fool you; it's a full chapter!)

## Session Info
```{r}
sessionInfo()
```





